Introducing the \NAME Framework\\ for Efficient Intrusion Detection\\ in the Internet of Things

### 2014-10-12 paper.tex
IEEE only accepts figure and table floats
added first draft background

ID algorithms targeting MANETs share a common design pattern as illustrated in
pseudocode listing \ref{alg:id-algo-pattern}. It consists of two functions:
\ttt{process\_message} and \ttt{do\_housekeeping}. The former function
represents the fact that ID algorithms inspect the network traffic that is
captured by the wireless radio and process it, looking for malicious activity.
The latter allows for algorithms to perform periodical tasks, checking their
internal information and optionally act upon it. Some algorithms will implement
both functions, some don't, but the abstract structure is applicable for a
significant number of algorithms.

\begin{listing}
\begin{algorithmic}[1]
  \Require{nodes, global storage for information about nodes}
  \Function{process\_message}{$msg$}
    \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
     \State \dots \Comment{analyse byte-sequences}
    \EndFor
    \Let{$nodes_x$}{value}  \Comment{optionally update node info}
    \State \Call{send}{$nodes_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send}
  \EndFunction
  \State
  \Function{do\_housekeeping}{}
    \ForEach{$node \in nodes$} \label{alg:id-algo-pattern-loop2}
      \If{$node > \dots$} \Comment{validate recorded value}
        \State \dots \Comment{take actions}
      \EndIf
    \EndFor
  \EndFunction
  \State
\end{algorithmic}
\caption{ID Algorithm Pattern\label{alg:id-algo-pattern}}
\end{listing}

The parts of interest here the iterations at lines
\ref{alg:id-algo-pattern-loop1} and \ref{alg:id-algo-pattern-loop2}, and the
interaction with the network at line \ref{alg:id-algo-pattern-send}.

Theory of Loop Fusion \cite{darte2000complexity}

Examples of code/loop fusion (e.g. Haskell Stream Fusion) \cite{coutts2007stream}

Function Inlining (e.g. TinyOS) \cite{gay2007software,gay2003nesc}

>>>>

The underlying problem addressed in this paper is of an economic nature: adding
ID doesn't add functionality, so its impact should be reduced as much as
possible.

In literature, many ID algorithm are described, but hardly any are implemented.
If implementations would be available, one would typically apply a pattern as
shown in listing \ref{alg:id-algo-application}.

\begin{figure}[H]
\begin{algorithmic}[1]
  \Let{$msg$}{$network$} \Comment{all observed messages}
  \ForEach{$algorithm \in algorithms$}
    \State \Call{algorithm.progress\_message}{msg}
  \EndFor
  \State \dots
  \Comment{at a given interval}
  \ForEach{$algorithm \in algorithms$}
    \State \Call{algorithm.do\_housekeeping}{}
  \EndFor
\end{algorithmic}
\captionof{listing}{Sequential implementation of several ID algorithms\label{alg:id-algo-application}}
\end{figure}

The implementations would be linked and called in a sequential way, when new
messages are observed on the network and at a regular interval to allow them to
do internal housekeeping, processing the information extracted from the network
traffic, often in a statistical way.

This is a common design pattern of ID algorithms targetting MANETs. The
implementation of this pattern also shares common structures, as illustrated in
listing \ref{alg:id-algo-pattern}.

\begin{figure}[H]
\begin{algorithmic}[1]
  \Require{nodes, global storage for information about nodes}
  \Function{process\_message}{$msg$}
    \ForEach{$byte \in msg$} \label{alg:id-algo-pattern-loop1}
     \State \dots \Comment{analyse byte-sequences}
    \EndFor
    \Let{$nodes_x$}{value}  \Comment{optionally update node info}
    \State \Call{send}{$nodes_y$, ``info''} \Comment{optionally exchange info} \label{alg:id-algo-pattern-send1}
  \EndFunction
  \State
  \Function{do\_housekeeping}{}
    \ForEach{$node \in nodes$} \label{alg:id-algo-pattern-loop2} \label{alg:id-algo-pattern-common-data}
      \If{$node > \dots$} \Comment{validate recorded value}
        \State \dots \Comment{take actions}
        \State \Call{send}{$node$, ``info''} \Comment{e.g. communicate} \label{alg:id-algo-pattern-send2}
      \EndIf
    \EndFor
  \EndFunction
  \State
\end{algorithmic}
\captionof{listing}{ID Algorithm Pattern\label{alg:id-algo-pattern}}
\end{figure}

The parts of interest here are the loops at lines
\ref{alg:id-algo-pattern-loop1} and \ref{alg:id-algo-pattern-loop2}, the common
\emph{nodes} data on line \ref{alg:id-algo-pattern-common-data} and the
interaction with the network at lines \ref{alg:id-algo-pattern-send1} and
\ref{alg:id-algo-pattern-send2}.

\subsection{Loops and Common Data}

Given several ID algorithms, that all perform the same parsing loop, over the
same byte stream, performing comparable operations to find similar patterns,
one would clearly benefit from combining these loops, avoiding redundant
reevaluation of the same byte patterns.

The same goes for common data. Each algorithm not only has to parse the same
message, it also has to store information about nodes within its own scope,
resulting in a lot of common data that is duplicated in memory.

The only way to solve this, would be to change the provided implementation,
pulling up the loops over the message and the nodes to a level above the loop
over the algorithms. But this comes at a development cost. This cost typically
isn't a one-time cost, e.g. in case a new version of the algorithm is released,
or simply when a different set of algorithms is selected. Finally, changing, or
simply implementing the algorithms from scratch, also holds the risk of making
mistakes.

\subsection{Network Usage}

But, as said before, there are hardly any implementations readily available.
Probably the main reason for this is the heterogeneity of the IoT. A prime
example is the \ttt{SEND} function. Different software stacks will have
different application programming interfaces (APIs) for accessing the network.
Simple, macro-like solutions can overcome this, but the dispersion of the calls
to access the network is causing the wireless radio to be on a lot more that
one would hope for.

A valid solution would try to combine messages in a single, reducing the access
to the network and even sharing some network overhead. The impact to implement
this, again would weigh to much on
